\bf Exploring probability distributions for bivariate temporal granularities

### 2020-02-07 main.tex
changes for asa journal rticle format

---
title: Exploring probability distributions for bivariate temporal granularities
authors:
- name: Sayani Gupta
  affiliation: Department of Econometrics and Business Statistics, Monash University
  email: Sayani.Gupta@monash.edu

- name: Rob J Hyndman
  affiliation: Department of Econometrics and Business Statistics, Monash University

- name: Dianne Cook
  affiliation: Department of Econometrics and Business Statistics, Monash University

- name: Antony Unwin
  affiliation: University of Augsburg

keywords:
- data visualization
- statistical distributions
- time granularities
- periodicties
- grammar of graphics
- R

abstract: |
 Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that need to be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic time granularities can be circular such as hour of the day, and day of the week, quasi-circular such as day of the month and aperiodic such as public or school holidays.

@@

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk]. 
bibliography: [bibliography.bib]
preamble: >
  \usepackage{mathtools}
  \usepackage{amssymb} 
  \usepackage{booktabs}
  \usepackage{longtable}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
library(tidyverse)
library(lubridate)
library(lvplot)
library(ggridges)
library(tsibble)
library(gravitas)
library(tsibble)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = "figure/", fig.align = "center", fig.show = "hold",
  cache = FALSE, cache.path = "cache/",
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
knitr::opts_knit$set(root.dir = here::here())
```

```{r external, include = FALSE}
# read_chunk('scripts/main.R')
```

```{r load}

```

# Introduction

<!--temporal granularities and why should we care -->

<!-- problem of analyzing data with such finer scales and classical ways to deal with it -->

Temporal data are available at various resolutions depending on the context. Social and economic data like GDP is often collected and reported at coarse temporal scales like monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded at much finer temporal scales. Energy consumption is collected every half an hour, while energy supply is collected every minute and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities [@aigner2011visualization].
It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to iteratively explore all possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

<!-- # existing ways and challenges -->

Calendar-based graphics [@wang2018calendar] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days. Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on the arrangement (linear vs. cyclic) or hierarchical order of the calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in @Bettini1998-ed. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular can be hour of the day, day of the week, that of quasi-circular can be day of the month and examples of aperiodic granularities can be public or school holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

<!--motivation -->

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates (see @2012-la) of usage across households, total kWh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data can be benefited from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. Although, the motivation came through the smart meter example, this is a problem which relates to any data that needs to be analyzed for different periodicities.

<!-- at finer or coarser scales can be benefited from the approach of Exploratory Data Analysis (EDA). EDA calls for utilizing visualization and transformation to explore data systematically. It is a process of generating hypothesis, testing them and consequently refining them through investigations. -->

<!--existing closely related work and how ours relate --> 
 <!-- This paper utilizes the nestedness of time granularities to obtain multiple-order-up granularities from single-order-up ones. -->

<!--actions on multiple order up - usage -->

<!-- Finally, visualizing data across single/multiple order-up granularities help us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. However, this approach can lead to a myriad of choices all of which are not useful. Analysts are expected to iteratively visualize these choices for exploring possible patterns in the data. But too many choices might leave him bewildered.  -->

<!-- challenges and how you are dealing wih it --> 
This work provides tools for systematically exploring bivariate granularities within the tidy workflow through the following:

<!-- Pairs of granularities are categorized as either a *harmony* or *clash*, where harmonies are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. -->

  * Formal characterization of cyclic granularities

  * Facilitate manipulation of single and multiple order-up time granularities through cyclic calendar algebra

  * Checking feasibility of creating plots or drawing inferences for any two cyclic granularities

  * Recommend prospective probability distributions for exploring distributions of univariate dependent variable across pair of granularities

  <!-- Pairs of granularities can be categorized as either a *harmony* or *clash*, where harmonies are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. -->

@@

# Linear time granularities {#sec:linear-time}

<!-- general calendar categorization -->
<!-- Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Classes of time abstractions may be based on arrangement or hierarchical order of time, discussed in further details in \autoref{sec:arrangement} and \autoref{sec:order}. -->

Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Time granularities are **linear** if they respect the linear progression of time. Examples include hours, days, weeks and months.

## Definitions and relationships

<!-- Arrangement: Linear vs. Circular vs. aperiodic {#sec:arrangement} -->

<!-- The arrangement of the time domain can result in deconstructing time in linear, circular or aperiodic (nearly circular) ways.  -->

 There has been several attempts to provide the framework for formally characterizing time-granularities. One of the first attempts occur in @Bettini1998-ed with the help of the following definitions:

<!-- <!-- Representation with tsibble -->
<!-- Suppose we have a tsibble with a time index in one column and keys and variables in other columns. A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of row numbers (the index set) to the time index. A linear granularity is a mapping of row numbers to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. -->

<!-- <!-- What is calendar categorization --> 
<!-- What we need to add to this are additional categorizations of time that are not linear granularities and are useful to represent periodicity. Examples include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc. Many of these are circular, such as day-of-week, time-of-day. Some are nearly circular such as day-of-month. Some are irregular such as working day/non-working day. Let's call all of these "calendar categorizations". Anything that maps a time index to a categorical variable can be considered a **calendar categorization**.  -->

<!-- What is circular granularity and some examples -->
<!-- We specify the circular categorizations using modular arithmetic and call these **circular granularities**. The number of categories is essentially the periodicity of a circular time granularity. For example, suppose the time index is in minutes, and let $n_i$ be the number of categories created by the circular granularity $C_i$. Then the following categorizations can be computed. -->

<!-- In the following sections, we provide a formalism to all of these abstractions and identify their relationships and symbolic representations.  -->

<!-- ### Linear {#sec:linear-gran-def} -->

@@

**Discussion:** The first condition in Definition \ref{def:linear} implies that the granules in a linear granularity are non-overlapping and their index order is same as time order.
\autoref{fig:linear-time} shows the implication of this condition. It shows the linear granularities hours, days, weeks, months and years, each of which are unidirectional in nature and arranged from past to future. If we consider the chronon (see @aigner2011visualization) as hourly, the time domain with T hours will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on. Each granule of the linear granularities shown are represented through a box.

```{r linear-time,echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear granularities"}
knitr::include_graphics("Figs/linear-arr.png")
```

[@Bettini1998-ed] talks about the relationships of linear time granularities and structure of a calendar and also relates them to the notion of periodicity in time.

\begin{definition}\label{def:finerthan}
(finer than): A linear granularity G is finer than a  linear granularity H, denoted $G \preceq H$, if
for each index i, there exists an index j such that $G(i) \subset H(j)$.
\end{definition}.

@@

**Discussion:** $G \trianglelefteq H$ if each granule H(j) is the union of some granules of G.  $G \preceq H$ if every granule in G is a subset of some granule in H. In \autoref{fig:groups-finer}, both $day \trianglelefteq week$ and $day \preceq week$ hold true. However, weekends are finer than weeks but do not group into weeks. And weekends neither groups into or finer than public holidays. Consider another example where $day \trianglelefteq month$. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions, .

```{r groups-finer,echo=FALSE,out.width = "100%",fig.cap="Illustrations of groups into and finer than relationships. Weekends are finer than weeks but do not group into weeks. And weekends neither groups into or finer than public holidays"}
knitr::include_graphics("Figs/groupsandfiner.png")
```

@@

**Discussion:** A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, ${S_0,...,S_{R-1}}$ describing the granules of H within one period;
(ii) the value of P; 
(iii) the indexes of first and last granules in H, if their value is not infinite. 
Then, if ${S_0,...,S_{R-1}}$ are the sets of indexes of G describing
$H (0), . . . , H (R - 1)$, respectively, then the description of an arbitrary granule H(j) is given by: $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$. Also, granularities can be periodical with respect to other granularities, except for a finite number of spans of time where they behave in an anomalous way [@Bettini2000-vy].

@@

**Discussion:** Intuitively, the definition requires that all granules of H within the span of time between two exceptions have the same periodical behavior, characterized by $R$ and $P$.

@@

**Discussion:** Given the set of all granularities defined over the time domain $(Z; <)$, and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

<!-- Most practical problems seem to require only a granularity system containing a set of time granularities which are all periodical (or quasi-periodical) with respect to a basic granularity. The assumption is there should be one bottom granularity and all the other granularities are generated from the bottom granularity by calendar algebraic operations. -->

## Computation through calendar algebra

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a “bottom” granularity and calendar algebra operations are designed to generate
new granularities from the bottom one or recursively, from those already generated.
The calendar algebra consists of two kinds of operations: grouping-oriented and
granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. Example can be to consider a calendar with only two linear granularities "minute" and "hour" and "hour" is generated by grouping every $60$ "minutes". The granule-oriented operations do not change the granules of a granularity, but rather make choices of which granules should remain in the new granularity. For example, one can choose to look at the granularity "Monday" and hence select only "Monday" from linear granularity "day".

@@

- **The grouping operation** : Let $G1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation $Group_m(G)$ generates a new granularity $G2$, by partitioning the granules of $G$ into m-granule groups and making each group a granule of the resulting granularity.
More precisely, $G2 = Group_m(G1)$ is the full-integer labeled granularity such that for
each integer $i$, $G2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G1(j)$.

**Examples** :

  - $minute = Group_{60}(second)$
  - $hour = Group_{60}(minute)$,
  - $day = Group_{24}(hour)$,
  - $week = Group_{7}(day)$,

@@

- **The altering-tick operation** : Let G1, G2 be full-integer labeled granularities, and $l$, $k$, $m$ integers, where G2 partitions G1, and $1 \leq l \leq m$. The altering-tick operation $Alter^{m}_{l,k} (G2, G1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the $l$th granule of each group has $|k|$ additional (or fewer when k < 0) granules of G2.

**Example** :

@@

, where the granularity pseudomonth is generated by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days. [@Ning2002-tf]

For more variations of grouping operations and granule oriented operations, the readers are recommended to see @Ning2002-tf.

# Cyclic time granularities {#sec:cyclic-gran}

<!-- sub: regular and irregular mapping leads to circular and quasi-circular granularities -->

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and key is a set of variables that define observational units over time. In a tsibble each observation should be uniquely identified by index and key.

A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble. Lining up with the the assumption in @Bettini2000-vy, that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here
\usepackage{mathtools} \usepackage{amssymb}  \usepackage{booktabs} \usepackage{longtable}

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Exploring probability distributions for bivariate temporal granularities}

  \author{
        Sayani Gupta \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Rob J Hyndman \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Dianne Cook \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Antony Unwin \\
    University of Augsburg\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Exploring probability distributions for bivariate temporal granularities}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that need to be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic time granularities can be circular such as hour of the day, and day of the week, quasi-circular such as day of the month and aperiodic such as public or school holidays.

@@

Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 \citep{Wickham2009pk}.
\end{abstract}

\noindent%
{\it Keywords:} data visualization, statistical distributions, time granularities, periodicties, grammar of graphics, R
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Temporal data are available at various resolutions depending on the context. Social and economic data like GDP is often collected and reported at coarse temporal scales like monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded at much finer temporal scales. Energy consumption is collected every half an hour, while energy supply is collected every minute and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities \citep{aigner2011visualization}.
It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA \citep{Tukey1977-jx} which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to iteratively explore all possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

Calendar-based graphics \citep{wang2018calendar} are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days. Any sub-daily resolution temporal data can also be displayed using this type of faceting \citep{Wickham2009pk} with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on the arrangement (linear vs.~cyclic) or hierarchical order of the calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in \citet{Bettini1998-ed}. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular can be hour of the day, day of the week, that of quasi-circular can be day of the month and examples of aperiodic granularities can be public or school holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate \citep{G_Grolemund2011-vm} creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates (see \citet{2012-la}) of usage across households, total kWh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data can be benefited from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. Although, the motivation came through the smart meter example, this is a problem which relates to any data that needs to be analyzed for different periodicities.

This work provides tools for systematically exploring bivariate granularities within the tidy workflow through the following:

\begin{itemize}
\item
  Formal characterization of cyclic granularities
\item
  Facilitate manipulation of single and multiple order-up time granularities through cyclic calendar algebra
\item
  Checking feasibility of creating plots or drawing inferences for any two cyclic granularities
\item
  Recommend prospective probability distributions for exploring distributions of univariate dependent variable across pair of granularities
\end{itemize}

@@

\hypertarget{sec:linear-time}{%
\section{Linear time granularities}\label{sec:linear-time}}

Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities \citep{aigner2011visualization}. Time granularities are \textbf{linear} if they respect the linear progression of time. Examples include hours, days, weeks and months.

\hypertarget{definitions-and-relationships}{%
\subsection{Definitions and relationships}\label{definitions-and-relationships}}

There has been several attempts to provide the framework for formally characterizing time-granularities. One of the first attempts occur in \citet{Bettini1998-ed} with the help of the following definitions:

@@

\textbf{Discussion:} The first condition in Definition \ref{def:linear} implies that the granules in a linear granularity are non-overlapping and their index order is same as time order.
\autoref{fig:linear-time} shows the implication of this condition. It shows the linear granularities hours, days, weeks, months and years, each of which are unidirectional in nature and arranged from past to future. If we consider the chronon (see \citet{aigner2011visualization}) as hourly, the time domain with T hours will have \(\lfloor T/24\rfloor\) days, \(\lfloor T/(24*7)\rfloor\) weeks and so on. Each granule of the linear granularities shown are represented through a box.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{Figs/linear-arr}

}

\caption{The time domain distributed as linear granularities}\label{fig:linear-time}
\end{figure}

\citep{Bettini1998-ed} talks about the relationships of linear time granularities and structure of a calendar and also relates them to the notion of periodicity in time.

\begin{definition}\label{def:finerthan}
(finer than): A linear granularity G is finer than a  linear granularity H, denoted $G \preceq H$, if
for each index i, there exists an index j such that $G(i) \subset H(j)$.
\end{definition}

.

@@

\textbf{Discussion:} \(G \trianglelefteq H\) if each granule H(j) is the union of some granules of G. \(G \preceq H\) if every granule in G is a subset of some granule in H. In \autoref{fig:groups-finer}, both \(day \trianglelefteq week\) and \(day \preceq week\) hold true. However, weekends are finer than weeks but do not group into weeks. And weekends neither groups into or finer than public holidays. Consider another example where \(day \trianglelefteq month\). This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions, .

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{Figs/groupsandfiner}

}

\caption{Illustrations of groups into and finer than relationships. Weekends are finer than weeks but do not group into weeks. And weekends neither groups into or finer than public holidays}\label{fig:groups-finer}
\end{figure}

@@

\textbf{Discussion:} A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, \({S_0,...,S_{R-1}}\) describing the granules of H within one period;
(ii) the value of P;
(iii) the indexes of first and last granules in H, if their value is not infinite.
Then, if \({S_0,...,S_{R-1}}\) are the sets of indexes of G describing
\(H (0), . . . , H (R - 1)\), respectively, then the description of an arbitrary granule H(j) is given by: \(\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)\). Also, granularities can be periodical with respect to other granularities, except for a finite number of spans of time where they behave in an anomalous way \citep{Bettini2000-vy}.

@@

\textbf{Discussion:} Intuitively, the definition requires that all granules of H within the span of time between two exceptions have the same periodical behavior, characterized by \(R\) and \(P\).

@@

\textbf{Discussion:} Given the set of all granularities defined over the time domain \((Z; <)\), and the granularity relationship \(\trianglelefteq\) (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to \(\trianglelefteq\).

\hypertarget{computation-through-calendar-algebra}{%
\subsection{Computation through calendar algebra}\label{computation-through-calendar-algebra}}

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra \citep{Ning2002-tf}. It is assumed that there exists a ``bottom'' granularity and calendar algebra operations are designed to generate
new granularities from the bottom one or recursively, from those already generated.
The calendar algebra consists of two kinds of operations: grouping-oriented and
granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. Example can be to consider a calendar with only two linear granularities ``minute'' and ``hour'' and ``hour'' is generated by grouping every \(60\) ``minutes''. The granule-oriented operations do not change the granules of a granularity, but rather make choices of which granules should remain in the new granularity. For example, one can choose to look at the granularity ``Monday'' and hence select only ``Monday'' from linear granularity ``day''.

@@

\begin{itemize}
\tightlist
\item
  \textbf{The grouping operation} : Let \(G1\) be a full-integer labeled granularity, and \(m\) a positive integer. The grouping operation \(Group_m(G)\) generates a new granularity \(G2\), by partitioning the granules of \(G\) into m-granule groups and making each group a granule of the resulting granularity.
  More precisely, \(G2 = Group_m(G1)\) is the full-integer labeled granularity such that for
  each integer \(i\), \(G2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G1(j)\).
\end{itemize}

\textbf{Examples} :

\begin{itemize}
\tightlist
\item
  \(minute = Group_{60}(second)\)
\item
  \(hour = Group_{60}(minute)\),
\item
  \(day = Group_{24}(hour)\),
\item
  \(week = Group_{7}(day)\),
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \textbf{The altering-tick operation} : Let G1, G2 be full-integer labeled granularities, and \(l\), \(k\), \(m\) integers, where G2 partitions G1, and \(1 \leq l \leq m\). The altering-tick operation \(Alter^{m}_{l,k} (G2, G1)\) generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the \(l\)th granule of each group has \(|k|\) additional (or fewer when k \textless{} 0) granules of G2.
\end{itemize}

\textbf{Example} :

@@

, where the granularity pseudomonth is generated by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days. \citep{Ning2002-tf}

For more variations of grouping operations and granule oriented operations, the readers are recommended to see \citet{Ning2002-tf}.

\hypertarget{sec:cyclic-gran}{%
\section{Cyclic time granularities}\label{sec:cyclic-gran}}

We propose a formalism of cyclic time granularities through the tsibble \citep{wang2019tsibble} framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and key is a set of variables that define observational units over time. In a tsibble each observation should be uniquely identified by index and key.

A time domain, as defined by \citet{Bettini1998-ed}, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble. Lining up with the the assumption in \citet{Bettini2000-vy}, that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.
